### 2018-07-17 chapters/03-translator/forward-trans.tex
General structure (without 'background')

\section{The Forward Translation}\label{sec:forward-translation}
\glsreset{teo}
The translation process starts with semantic \LaTeX{} expressions. These expressions get analyzed and parsed by the \gls{mlp}, which produces a \gls{mlp-pt} of the input expression with additional information about each symbol. An abstract translator class analyzes each node and delegates them to specialized subtranslators. In this process an object called \gls{teo} is build. This \gls{teo} is needed to rebuild a string representation of the mathematical expression given by the \gls{mlp-pt}.

@@

Algorithm~\ref{alg:translation} is a simplified version of the translator process. The main progress happens in \cref{line:transLeaf,line:transSeq}. There are several cases what $r$ can be. Table~\ref{tab:allTypesTable} on page \pageref{tab:allTypesTable} gives an overview of all different types in a \gls{mlp-pt}. A more detailed explanation of the types can be found in~\cite{POM-Tagger}.

@@

Another more questionable translation decision are alphanumerical expressions. As explained in table~\ref{tab:allTypesTable}, the \gls{mlp} handles strings of letters and numbers differently, depending on the order of the symbols. The reason is, that an expression such as '$4b$' is usually considered to be a multiplication of $4$ and '$b$', while '$b4$' looks like indexing '$b$' by $4$. While the first example produces two nodes, namely $4$ and '$b$', the second example '$b4$' produces just a single alphanumerical node in the \gls{mlp-pt}. The first expression would be translated as a multiplication of $4$ and '$b$'. Because '$4b$' looks like '$4 \cdot b$' and the multiplication is commutative, we would assume that '$4b$' and '$b4$' are equivalent. This is one reason why we interpret such alphanumerical expressions as multiplications of the symbols. On the other hand, an expression such as '$energy$', would be an alphanumerical node as well and could be interpreted as a single variable named 'energy'. In that case, someone could wonder why the variable '$energy$' was translated as '$e \cdot n \cdot e \cdot r \cdot g \cdot y$'. While in physics or engineering variables possibly appear with longer names, such as the 'energy' example, it is more common to use single symbols for variables in mathematics~\cite{Notation:History}. Therefore, we interpret such alphanumerical expressions as multiplications of variables rather than one variable with an alphanumerical name.

>>>>

\section{The Forward Translation}\label{sec:forward-translation}
%\glsreset{teo}
The translation process starts with semantic \LaTeX{} expressions. These expressions get analyzed and parsed by the \gls{mlp}, which produces a \gls{mlp-pt} of the input expression with additional information about each symbol. An abstract translator class analyzes each node and delegates them to specialized subtranslators. In this process an object called \gls{teo} is build. This \gls{teo} is needed to rebuild a string representation of the mathematical expression given by the \gls{mlp-pt}.

@@

Algorithm~\ref{alg:translation} is a simplified version of the translator process. The main progress happens in \cref{line:transLeaf,line:transSeq}. There are several cases what $r$ can be. Table~\ref{tab:allTypesTable} on page \pageref{tab:allTypesTable} gives an overview of all different types in a \gls{mlp-pt}. A more detailed explanation of the types can be found in~\parencite{POM-Tagger}.

@@

Another more questionable translation decision are alphanumerical expressions. As explained in table~\ref{tab:allTypesTable}, the \gls{mlp} handles strings of letters and numbers differently, depending on the order of the symbols. The reason is, that an expression such as '$4b$' is usually considered to be a multiplication of $4$ and '$b$', while '$b4$' looks like indexing '$b$' by $4$. While the first example produces two nodes, namely $4$ and '$b$', the second example '$b4$' produces just a single alphanumerical node in the \gls{mlp-pt}. The first expression would be translated as a multiplication of $4$ and '$b$'. Because '$4b$' looks like '$4 \cdot b$' and the multiplication is commutative, we would assume that '$4b$' and '$b4$' are equivalent. This is one reason why we interpret such alphanumerical expressions as multiplications of the symbols. On the other hand, an expression such as '$energy$', would be an alphanumerical node as well and could be interpreted as a single variable named 'energy'. In that case, someone could wonder why the variable '$energy$' was translated as '$e \cdot n \cdot e \cdot r \cdot g \cdot y$'. While in physics or engineering variables possibly appear with longer names, such as the 'energy' example, it is more common to use single symbols for variables in mathematics~\parencite{Notation:History}. Therefore, we interpret such alphanumerical expressions as multiplications of variables rather than one variable with an alphanumerical name.

### 2018-07-17 chapters/03-translator/gen-app.tex
General structure (without 'background')

The overall goal of the whole project was to provide an interactive one click-translation for each formula in the \gls{drmf} to a specified \gls{cas}. The program mainly supports translations from and to the \gls{cas} \Maple. However, the program is designed to support other \gls{cas}. In the current stage it also supports translations to \Mathematica{}, but the backend library needs to get extended for a full support.

In this chapter, the functionality and the design of the translator will be explained. The structure follows the development process and explains our decisions, problems and solutions. It starts with an overview of our approaches in~\ref{sec:gen-appr} and solutions for the problems. We will also focus on the theory of our backend libraries in~\ref{subsec:libs} and how to extend them in~\ref{subsec:extension}. The sections~\ref{sec:forward-translation} and \ref{sec:backward-translation} explains the forward and backward translator separately.

\section{General Approach and Goals}\label{sec:gen-appr}
Initially this project aimed for translations not only for semantic \LaTeX{} expressions but for generic \LaTeX{} as well. However, as discussed before, the semantic information in generic \LaTeX{} is hardly accessible (as seen in \ref{subsec:semantic-latex}). Supporting generic \LaTeX{} translations would go beyond the scope of a Master's thesis. Hence, we focused on translations for semantic \LaTeX{} expressions. Furthermore, the translator should support multiple \gls{cas} and because of the work in progress state of the \Macro{} it should be easily extendable and changeable. Due to the time limitations of the project we focused our implementations on the support for \Maple{} and \Mathematica.

Another strive for a goal was to keep a translation as straight forward as possible. In other words, the structure of the translated expression should be similar to the structure of the input expression. This makes it easier for a user to follow the translation process and debug the program. This intention should not to be confused with the task for \textit{appropriate} translations. We will see that a translation process (especially the backward translation) can easily change expressions without changing the semantics. We want to avoid that behavior in our translations as often as possible.

Furthermore, the translation process should solve problems of existing technologies, presented in section~\ref{sec:related-work}. In the following subsection, we will focus on known problems and present solutions before we start to explain implementation details.

\subsection{Translation Problems}
Most of the existing translation tools cannot preserve the semantic information during the translation process (see section~\ref{sec:related-work}). Therefore, our program should not lose semantic information during the process. However, the semantics can be changed during a translation. This happens when a definition or property of a mathematical object varies from system to system. Our reference systems for all mathematical \LaTeX{} expressions are the \gls{dlmf} and the \gls{drmf}.

Let us focus on translations of mathematical objects that are not defined in one of the languages, first. Obviously, we cannot translate these objects \textit{directly}. Our workaround is to translate not the mathematical object itself, but the definition of the object. For example, the \textit{Gudermannian}~\cite[(4.23.10)]{NIST:DLMF} $\Gudermannian@{x}$ function can be defined by
\begin{equation}\label{eq:gudermannian}
\Gudermannian@{x} := \atan@{\sinh@{x}} \quad x \in \Real.
\end{equation}
This function is not implemented in \Maple. Therefore, we translate its definition~\eqref{eq:gudermannian}
\begin{equation}\label{eq:guder-trans}
\verb|\Gudermannian@{x}| \overset{\langMaple}{\mapsto} \texttt{arctan(sinh(x))}.
\end{equation}

@@

Instead of none possible translation, how can we handle multiple alternative translations? We already discussed the problem of the inverse trigonometric function $\acot@{x}$. Because of a different branch cut in \Maple{} (compared to the \gls{dlmf} definition), there are alternative translations available (see~\eqref{eq:acot-alternatives}). In such cases, we define the most intuitive, the direct translation. In addition, we inform the user about the available alternative translations and the branch cut issues. Hence,
\begin{equation}
\verb|\acot@{x}| \overset{\langMaple}{\mapsto} \texttt{arccot(x)}.
\end{equation}

@@

Note that \Maple{} also uses the symbol \verb|$| in few functions. This can cause trouble in the replacement process. However, we can handle all of the few cases by adding addition parenthesis. Consider the differentiation function in \Maple
\begin{equation}
\verb|diff(f, [x$n])|,
\end{equation}
with $f$ as an algebraic expression or an equation, $x$ the name of the differentiation variable and $n$ for the $n$-th order differentiation. Therefore
\begin{equation}
\deriv[2]{x^2}{x} = \verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|.
\end{equation}
But the translation pattern in \Maple{} would be
\begin{equation}
\verb|diff($1, [$2$$0])|.
\end{equation}
Replace the placeholders sequentially by the arguments ends in
\begin{eqnarray*}
&&\verb|diff($1, [$2$$0])|\\
\text{Replace index 0 by $2$\hphantom{$^2$}}& \rightarrow &\verb|diff($1, [$2$2])|\\
\text{Replace index 1 by $x^2$}& \rightarrow &\verb|diff(x^2, [$2$2])|\\
\text{Replace index 2 by $x$\hphantom{$^2$}}& \rightarrow &\verb|diff(x^2, [xx])|.
\end{eqnarray*}
We use parenthesis to solve this issue and define the translation pattern with
\begin{equation}
\verb|diff($1, [$2$($0)])|.
\end{equation}

>>>>

%The overall goal of the whole project was to provide an interactive one click-translation for each formula in the \gls{drmf} to a specified \gls{cas}. The program mainly supports translations from and to the \gls{cas} \Maple. However, the program is designed to support other \gls{cas}. In the current stage it also supports translations to \Mathematica{}, but the backend library needs to get extended for a full support.

%In this chapter, the functionality and the design of the translator will be explained. The structure follows the development process and explains our decisions, problems and solutions. It starts with an overview of our approaches in~\ref{sec:gen-appr} and solutions for the problems. We will also focus on the theory of our backend libraries in~\ref{subsec:libs} and how to extend them in~\ref{subsec:extension}. The sections~\ref{sec:forward-translation} and \ref{sec:backward-translation} explains the forward and backward translator separately.

%\section{General Approach and Goals}\label{sec:gen-appr}
%Initially this project aimed for translations not only for semantic \LaTeX{} expressions but for generic \LaTeX{} as well. However, as discussed before, the semantic information in generic \LaTeX{} is hardly accessible (as seen in \ref{subsec:semantic-latex}). Supporting generic \LaTeX{} translations would go beyond the scope of a Master's thesis. Hence, we focused on translations for semantic \LaTeX{} expressions. Furthermore, the translator should support multiple \gls{cas} and because of the work in progress state of the \Macro{} it should be easily extendable and changeable. Due to the time limitations of the project we focused our implementations on the support for \Maple{} and \Mathematica.

%Another strive for a goal was to keep a translation as straight forward as possible. In other words, the structure of the translated expression should be similar to the structure of the input expression. This makes it easier for a user to follow the translation process and debug the program. This intention should not to be confused with the task for \textit{appropriate} translations. We will see that a translation process (especially the backward translation) can easily change expressions without changing the semantics. We want to avoid that behavior in our translations as often as possible.

%Furthermore, the translation process should solve problems of existing technologies, presented in section~\ref{sec:related-work}. In the following subsection, we will focus on known problems and present solutions before we start to explain implementation details.

\section{Translation Problems}
Most of the existing translation tools cannot preserve the semantic information during the translation process (see section~\ref{sec:related-work}). Therefore, our program should not lose semantic information during the process. However, the semantics can be changed during a translation. This happens when a definition or property of a mathematical object varies from system to system. Our reference systems for all mathematical \LaTeX{} expressions are the \gls{dlmf} and the \gls{drmf}.

Let us focus on translations of mathematical objects that are not defined in one of the languages, first. Obviously, we cannot translate these objects \textit{directly}. Our workaround is to translate not the mathematical object itself, but the definition of the object. For example, the \textit{Gudermannian}~\parencite[(4.23.10)]{NIST:DLMF} $\Gudermannian{x}$ function can be defined by
\begin{equation}\label{eq:gudermannian}
\Gudermannian{x} := \atan{\sinh{x}} \quad x \in \Real.
\end{equation}
This function is not implemented in \Maple. Therefore, we translate its definition~\eqref{eq:gudermannian}
\begin{equation}\label{eq:guder-trans}
\verb|\Gudermannian{x}| \overset{\langMaple}{\mapsto} \texttt{arctan(sinh(x))}.
\end{equation}

@@

Instead of none possible translation, how can we handle multiple alternative translations? We already discussed the problem of the inverse trigonometric function $\acot@{x}$. Because of a different branch cut in \Maple{} (compared to the \gls{dlmf} definition), there are alternative translations available (see~\eqref{eq:acot-alternatives}). In such cases, we define the most intuitive, the direct translation. In addition, we inform the user about the available alternative translations and the branch cut issues. Hence,
\begin{equation}
\verb|\acot{x}| \overset{\langMaple}{\mapsto} \texttt{arccot(x)}.
\end{equation}

@@

Note that \Maple{} also uses the symbol \verb|$| in few functions. This can cause trouble in the replacement process. However, we can handle all of the few cases by adding additional parenthesis. Consider the differentiation function in \Maple
\begin{equation}
\verb|diff(f, [x$n])|,
\end{equation}
with $f$ as an algebraic expression or an equation, $x$ the name of the differentiation variable and $n$ for the $n$-th order differentiation. Therefore
\begin{equation}
\frac{d^2x^2}{dx^2} = \verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|.
\end{equation}
But the translation pattern in \Maple{} would be
\begin{equation}
\verb|diff($1, [$2$$0])|.
\end{equation}
Replace the placeholders sequentially by the arguments ends in
\begin{eqnarray*}
&&\verb|diff($1, [$2$$0])|\\
\text{Replace index 0 by $2$\hphantom{$^2$}}& \rightarrow &\verb|diff($1, [$2$2])|\\
\text{Replace index 1 by $x^2$}& \rightarrow &\verb|diff(x^2, [$2$2])|\\
\text{Replace index 2 by $x$\hphantom{$^2$}}& \rightarrow &\verb|diff(x^2, [xx])|.
\end{eqnarray*}
We use parenthesis to solve this issue and define the translation pattern with
\begin{equation}
\verb|diff($1, [$2$($0)])|.
\end{equation}

### 2018-07-17 chapters/04-evaluation/evaluation.tex
General structure (without 'background')

First of all, each semantic \LaTeX{} macro ties a specific character sequence to a well-defined mathematical object. This definition can be found at the \gls{dlmf} or \gls{drmf}. Therefore, our reference source for semantic \LaTeX{} expressions is always the \gls{dlmf} or \gls{drmf}. This gives us the opportunity to compare more than the string representations of the mathematical objects in a translation process. For example, consider the simple translation of the cosine function
\begin{equation}\label{eq:simple-cosine}
\verb|\cos@{x}| \overset{\langMaple}{\mapsto} \texttt{cos(x)}.
\end{equation}
Our verification and validation techniques, to proof for example that~(\ref{eq:simple-cosine}) is an \textit{appropriate} translation, can be grouped into three parts. We present in the following sections these three parts and conclusively, we use all these techniques to actually verify translations for a test suite directly extracted from the \gls{dlmf}.

@@

There is currently only one exception known where a round trip test does not reach a fixed point representation at all: Legendre's incomplete elliptic integrals~\cite[(19.2.4-7)]{NIST:DLMF} are defined with the amplitude $\phi$ in the first argument in the \DLMF, while \Maple{} takes the trigonometric sine of the amplitude as the first argument. Therefore, the forward and backward translations are defined as
\begin{eqnarray}
\verb|\EllIntF@{\phi}{k}| & \overset{\langMaple}{\mapsto} & \verb|EllipticF(sin(phi),k)|,\\
\verb|\EllIntF@{\asin@{\phi}}{k}| & \overset{\langMaple}{\mapsfrom} & \verb|EllipticF(phi,k)|.
\end{eqnarray}

@@

With this technique we can detect translation errors such as in (\ref{eq:wrong-trans-1} and \ref{eq:wrong-trans-2}). Consider the \gls{dlmf} equation for the sine and cosine function~\cite[(4.21.2)]{NIST:DLMF}
\begin{equation}
\sin@{u+v} = \sin@@{u}\cos@@{v} + \cos@@{u}\sin@@{v}.
\end{equation}
Assume the translator would forward translate the expression based on (\ref{eq:wrong-trans-1},~\ref{eq:wrong-trans-2}). Than
\begin{eqnarray}
\verb|\sin@{u + v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u + v)|,\\
\verb|\sin@@{u}\cos@@{v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u)*sin(v)|,\\
\verb|\cos@@{u}\sin@@{v}| & \overset{\langMaple}{\mapsto} & \verb|sin(u)*cos(v)|.
\end{eqnarray}
This produces the equation in \Maple
\begin{equation}
\cos@{u+v} = \cos@{u}\sin@{v} + \sin@{u}\cos@{v},
\end{equation}
which is wrong. Since the expression is correct before the translation, we conclude an error during the translation process.

However, there are two essential problems with this approach. Testing the mathematical equivalence of expressions is hard to solve and \gls{cas} have trouble to test even simple equations. Furthermore, this approach only checks forward translations because there is no way to check equivalence of expressions in \LaTeX{} automatically (again this could become feasible with our translator). We use \Maple's \textit{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \textit{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. This simplification function is the most powerful function to check the equivalence in \Maple. However, there are several cases where the simplification fails. We implement some tricks to help \textit{simplify}: pre-conversion of the functions of the expression to other special functions is able to help \textit{simplify}, because the algorithms in the background can use different simplification methods for different functions. For example, the user can use the \Maple{} function \textit{convert} to convert the expression
\begin{equation}
\sinh@{x} + \sin@{x}
\end{equation}
to an equivalent representation using the exponential function:
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions we are able to improve the simplification process in \Maple. However, the limitations of the \textit{simplify} function are still the weakest part of this verification approach. Consider the equation~(\ref{eq:branch-cut-issue}) from \cref{subsec:branch-cut-issues}. 
\begin{equation}\label{eq:branch-cut-near}
\displaystyle \ParabolicU@{0}{z} = \sqrt{\frac{z}{2\cpi}}\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}
\end{equation}
As already discussed, the modified Bessel function of the second kind~\cite[(12.7.10)]{NIST:DLMF} has its branch cut in \Maple{} at $z < 0$. However, the argument of $K$ contains $z^2$. For now, just focus on
\begin{equation}\label{eq:besselK}
\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}
\end{equation}
Because of $z^2$, if $|\ph@{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of (\ref{eq:besselK}) would be no longer on the principal branch. However, \Maple{} will still compute (\ref{eq:besselK}) on the principal branch independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph@{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one has to use analytic continuation for the right-hand side of equation (\ref{eq:branch-cut-near}). In subsection~\ref{subsec:branch-cut-issues} we also shown that the complete right-hand side of equation~(\ref{eq:branch-cut-near}) has three branch cuts, caused by the square root function and the $z^2$ argument of the modified Bessel function of the second kind. Such complex examples cannot be simplified yet.

@@

For example, reconsider equation~(\ref{eq:branch-cut-near}) from \cite[(12.7.10)]{NIST:DLMF}. Define the difference of right-hand side and left-hand side
\begin{equation}\label{eq:difference}
D(z) := \ParabolicU@{0}{z} - \sqrt{\frac{z}{2\cpi}}\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2},
\end{equation}
and compute $D(z)$ for some values of $z$. Table~\ref{tab:computations-for-difference} presents four computations for $D(z)$, one value for each quadrant in the complex plane.
\begin{table}[ht]
\centering
\begin{tabular}{rcc}
	\hline
	$z\ \ $ & & $D(z)$\\
	\hline
	\tableRowSpace{} $1+\iunit$ & & $2 \cdot 10^{-10} - 2 \cdot 10^{-10} \iunit$\\
	$-1+\iunit$& & $2.222121916 - 1.116719816 \iunit$\\
	$-1-\iunit$& & $2.222121916 + 1.116719816 \iunit$\\
	$1-\iunit$ & & $2 \cdot 10^{-10} + 2 \cdot 10^{-10} \iunit$\\
	\hline
\end{tabular}
\caption{Four computations of $D(z)$ in \Maple.}
\label{tab:computations-for-difference}
\end{table}

@@

Two of the test cases (lines 42 and 43) contain macros we are not able to translate yet. Those macros are \verb|\subplus| and \verb|\CFK|, which are used for continued fractions. They have been created for another translation project~\cite[§5]{CICM:Paper} for continued fractions defined in the \textit{Handbook of Continued Fractions for Special Functions} (2008) \cite{CF:Handbook}. The conversion process for the \gls{cfsf} \cite{CF:Maple} dataset from \Maple{} and the \gls{ecf} \cite{CF:eCF,eCF:FinalReport} dataset from Wolfram uses special file types. Therefore, our translator is not able to handle those macros yet.

However, we translate 108 test cases and for 84 of them the simplification function of \Maple{} returns zero as the difference of the left-hand side and right-hand side. The 24 remaining cases are translated, but the simplification function is not able to simplify the difference to zero. Therefore, we try to improve this part. Line 46~\cite[(18.5.10)]{NIST:DLMF} for example, can be simplified to zero, if $n$ is previously set to an arbitrary integer value. For all those 24 test cases, we perform also manual numerical tests, which reveal an overall sign error in equation~\cite[(14.5.14)]{NIST:DLMF}\footnote{The equation had originally been stated as shown in equation~(\ref{eq:signerror}). The \gls{dlmf} has already updated the wrong equation.}
\begin{equation}\label{eq:signerror}
\displaystyle \LegendreP[1/2]{\nu}@{\cosh@@{\xi}} = - \left( \frac{2}{\cpi \sinh@@{\xi}} \right)^{1/2} \cosh@{\left( \nu + \tfrac{1}{2} \right) \xi}.
\end{equation}
Besides this wrong equation, all of these manual numerical tests return zero for the difference between the left-hand side and right-hand side. Of course it is still not possible to conclude the equivalence based on our manual numerical tests, as explained above.

@@

The translator cannot translate 1,933 test cases. First of all, 32 lines contain neither an equation nor a relation. Because we have created this dataset for our verification approaches, we do not translate those 32 lines at all. Most of the other failures are produced because of missing semantic information in the \LaTeX{} expression or the formulae contains a macro that cannot be translated in the current state of the program, such as in 978 lines that contain a macro that cannot be translated. For example, in 171 lines one of the symmetric elliptic integrals~\cite[(19.16.1-6)]{NIST:DLMF} appears and we do not provide translation patterns for those integrals, because there is no direct translation possible yet. A workaround would be to allow translations based on the definitions, see~(\ref{eq:gudermannian}).

>>>>

First of all, each semantic \LaTeX{} macro ties a specific character sequence to a well-defined mathematical object. This definition can be found at the \gls{dlmf} or \gls{drmf}. Therefore, our reference source for semantic \LaTeX{} expressions is always the \gls{dlmf} or \gls{drmf}. This gives us the opportunity to compare more than the string representations of the mathematical objects in a translation process. For example, consider the simple translation of the cosine function
\begin{equation}\label{eq:simple-cosine}
\verb|\cos{x}| \overset{\langMaple}{\mapsto} \texttt{cos(x)}.
\end{equation}
Our verification and validation techniques, to proof for example that~(\ref{eq:simple-cosine}) is an \textit{appropriate} translation, can be grouped into three parts. We present in the following sections these three parts and conclusively, we use all these techniques to actually verify translations for a test suite directly extracted from the \gls{dlmf}.

@@

There is currently only one exception known where a round trip test does not reach a fixed point representation at all: Legendre's incomplete elliptic integrals~\parencite[(19.2.4-7)]{NIST:DLMF} are defined with the amplitude $\phi$ in the first argument in the \DLMF, while \Maple{} takes the trigonometric sine of the amplitude as the first argument. Therefore, the forward and backward translations are defined as
\begin{eqnarray}
\verb|\EllIntF@{\phi}{k}| & \overset{\langMaple}{\mapsto} & \verb|EllipticF(sin(phi),k)|,\\
\verb|\EllIntF@{\asin@{\phi}}{k}| & \overset{\langMaple}{\mapsfrom} & \verb|EllipticF(phi,k)|.
\end{eqnarray}

@@

With this technique we can detect translation errors such as in (\ref{eq:wrong-trans-1} and \ref{eq:wrong-trans-2}). Consider the \gls{dlmf} equation for the sine and cosine function~\parencite[(4.21.2)]{NIST:DLMF}
\begin{equation}
\sin \left(u+v\right) = \sin{u}\cos{v} + \cos{u}\sin{v}.
\end{equation}
Assume the translator would forward translate the expression based on (\ref{eq:wrong-trans-1},~\ref{eq:wrong-trans-2}). Than
\begin{eqnarray}
\verb|\sin@{u + v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u + v)|,\\
\verb|\sin@@{u}\cos@@{v}| & \overset{\langMaple}{\mapsto} & \verb|cos(u)*sin(v)|,\\
\verb|\cos@@{u}\sin@@{v}| & \overset{\langMaple}{\mapsto} & \verb|sin(u)*cos(v)|.
\end{eqnarray}
This produces the equation in \Maple
\begin{equation}
\cos\left(u+v\right) = \cos{u}\sin{v} + \sin{u}\cos{v},
\end{equation}
which is wrong. Since the expression is correct before the translation, we conclude an error during the translation process.

However, there are two essential problems with this approach. Testing the mathematical equivalence of expressions is hard to solve and \gls{cas} have trouble to test even simple equations. Furthermore, this approach only checks forward translations because there is no way to check equivalence of expressions in \LaTeX{} automatically (again this could become feasible with our translator). We use \Maple's \textit{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \textit{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. This simplification function is the most powerful function to check the equivalence in \Maple. However, there are several cases where the simplification fails. We implement some tricks to help \textit{simplify}: pre-conversion of the functions of the expression to other special functions is able to help \textit{simplify}, because the algorithms in the background can use different simplification methods for different functions. For example, the user can use the \Maple{} function \textit{convert} to convert the expression
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an equivalent representation using the exponential function:
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions we are able to improve the simplification process in \Maple. However, the limitations of the \textit{simplify} function are still the weakest part of this verification approach. Consider the equation~(\ref{eq:branch-cut-issue}) from \cref{subsec:branch-cut-issues}. 
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right)
\end{equation}
As already discussed, the modified Bessel function of the second kind~\parencite[(12.7.10)]{NIST:DLMF} has its branch cut in \Maple{} at $z < 0$. However, the argument of $K$ contains $z^2$. For now, just focus on
\begin{equation}\label{eq:besselK}
K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right)
\end{equation}
Because of $z^2$, if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of (\ref{eq:besselK}) would be no longer on the principal branch. However, \Maple{} will still compute (\ref{eq:besselK}) on the principal branch independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one has to use analytic continuation for the right-hand side of equation (\ref{eq:branch-cut-near}). In subsection~\ref{subsec:branch-cut-issues} we also shown that the complete right-hand side of equation~(\ref{eq:branch-cut-near}) has three branch cuts, caused by the square root function and the $z^2$ argument of the modified Bessel function of the second kind. Such complex examples cannot be simplified yet.

@@

For example, reconsider equation~(\ref{eq:branch-cut-near}) from \parencite[(12.7.10)]{NIST:DLMF}. Define the difference of right-hand side and left-hand side
\begin{equation}\label{eq:difference}
D(z) := U(0,z) - \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
and compute $D(z)$ for some values of $z$. Table~\ref{tab:computations-for-difference} presents four computations for $D(z)$, one value for each quadrant in the complex plane.
\begin{table}[ht]
\centering
\begin{tabular}{rcc}
	\hline
	$z\ \ $ & & $D(z)$\\
	\hline
	\tableRowSpace{} $1+\iunit$ & & $2 \cdot 10^{-10} - 2 \cdot 10^{-10} \iunit$\\
	$-1+\iunit$& & $2.222121916 - 1.116719816 \iunit$\\
	$-1-\iunit$& & $2.222121916 + 1.116719816 \iunit$\\
	$1-\iunit$ & & $2 \cdot 10^{-10} + 2 \cdot 10^{-10} \iunit$\\
	\hline
\end{tabular}
\caption{Four computations of $D(z)$ in \Maple.}
\label{tab:computations-for-difference}
\end{table}

@@

Two of the test cases (lines 42 and 43) contain macros we are not able to translate yet. Those macros are \verb|\subplus| and \verb|\CFK|, which are used for continued fractions. They have been created for another translation project~\parencite[§5]{CICM:Paper} for continued fractions defined in the \textit{Handbook of Continued Fractions for Special Functions} (2008) \parencite{CF:Handbook}. The conversion process for the \gls{cfsf} \parencite{CF:Maple} dataset from \Maple{} and the \gls{ecf} \parencite{CF:eCF,eCF:FinalReport} dataset from Wolfram uses special file types. Therefore, our translator is not able to handle those macros yet.

However, we translate 108 test cases and for 84 of them the simplification function of \Maple{} returns zero as the difference of the left-hand side and right-hand side. The 24 remaining cases are translated, but the simplification function is not able to simplify the difference to zero. Therefore, we try to improve this part. Line 46~\parencite[(18.5.10)]{NIST:DLMF} for example, can be simplified to zero, if $n$ is previously set to an arbitrary integer value. For all those 24 test cases, we perform also manual numerical tests, which reveal an overall sign error in equation~\parencite[(14.5.14)]{NIST:DLMF}\footnote{The equation had originally been stated as shown in equation~(\ref{eq:signerror}). The \gls{dlmf} has already updated the wrong equation.}
\begin{equation}\label{eq:signerror}
\displaystyle P^{1/2}_{\nu}\left(\cosh{\xi}\right) = - \left( \frac{2}{\cpi \sinh{\xi}} \right)^{1/2} \cosh{\left( \nu + \tfrac{1}{2} \right) \xi}.
\end{equation}
Besides this wrong equation, all of these manual numerical tests return zero for the difference between the left-hand side and right-hand side. Of course it is still not possible to conclude the equivalence based on our manual numerical tests, as explained above.

@@

The translator cannot translate 1,933 test cases. First of all, 32 lines contain neither an equation nor a relation. Because we have created this dataset for our verification approaches, we do not translate those 32 lines at all. Most of the other failures are produced because of missing semantic information in the \LaTeX{} expression or the formulae contains a macro that cannot be translated in the current state of the program, such as in 978 lines that contain a macro that cannot be translated. For example, in 171 lines one of the symmetric elliptic integrals~\parencite[(19.16.1-6)]{NIST:DLMF} appears and we do not provide translation patterns for those integrals, because there is no direct translation possible yet. A workaround would be to allow translations based on the definitions, see~(\ref{eq:gudermannian}).
